---
title: "Which Group Uses City Shelter Predominantly? An Analyses of City Shelter Occupations"
author: Yi Fei Pang
date: today
date-format: long
abstract: "Shelter facilities are becoming more and more prevalent throughout the years. Based on the data from early-mid January of 2024, Mixed Adults seem to be the group that uses shelters the most followed by families and Youths use shelters the least. This suggests that there should be more resources in place that can help adults and families in these shelters, as well, maybe resources to help the youths so there will be less Mixed Adults in the shelters in the future."
format: pdf
toc: true
number-sections: true
bibliography: ref.bib
---

```{r}
#| message: false
#| echo: false


#### Workplace Setup ####

# Get required libraries
library(tidyverse)
library(knitr)
library(ggplot2)
library(ggrepel)
library(here)

# Read in necessary data
# cleaned_shelter <- read.csv(
#   file = here("outputs/data/shelter_cleaned.csv"),
#   show_col_type = FALSE
# )
```

## Introduction
As Canada experiences higher levels of immigration (Lavoie, 2023), global 
warming causing more natural disasters, and housing prices continue to 
rise (Connelly, 2022), many are left without a home at times. In the past few 
years, there has been multiple instances where shelters have been overrun 
because they were not prepared for the massive quantity of people in need.

As Canada reaches a historical high number of immigrants at almost 500,000
annually (statista, 2024) creating a situation where the people
coming to Canada in hopes of safety and a new life are greeted without a proper
home or shelter. This leaves hundreds of refugee families left on the streets.
(Lavoie, 2023)

Furthermore, as the housing crisis continues where home prices spike to 
abnormal heights, rent increases, and mortgage rates increase, many Canadians
are left without a home. This primarily affects young adults and youths who, 
in the past would be able to pay rent and buy homes with lower income levels.
(Pereira, 2022)

Since the need for shelters is becoming a national issue,
by analyzing the demographics of the people who use shelters along with the 
quantity of those demographics, cities can offer a more tailored solution to
provide for those who need shelters. Throughout my analysis, I will provide
statistics along with graphs to show findings for shelter users by demographics.

## Data
The data used in my analysis comes solely from OpenDataToronto.
Data was collected and analyzed using the statistical programming software R
[@r], with additional support from `tidyverse` [@rTidyverse], `ggplot2` 
[@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], 
`tibble` [@rTibble], `janitor` [@rJanitor], `ggrepel` [@rggrepel], and 
`here` [@rHere]. The collection, cleaning, modelling and analysis of the data
will be detailed later on in this paper.

## Shelter Occcupation
The dataset presents an extensive csv of people who use shelters, the detailed
location of the shelters, the capacity, and the demographics of the 
shelter-goers. The dates of this dataset ranges from the first day of January 
2024 up until January 21st.

## Data Cleaning
The data I acquired had a detailed description of the shelters. First, there
every row started with a date and is a variable required and that should be 
preserved through data cleaning. The dates ranged between 2024-01-01 all the way
to 2024-01-21. The next few columns were less important, these contained the 
Organization ID, name of the shelter, location address, postal code, city, and
province. The next few columns contain details about the program. The program 
ID, name, sector, model, area, and service type. One important column is the
sector because it identifies the demography of the people who use the shelter.
The rest of the data is not as useful for my analysis.

```{r}

```
## Similar Datasets
There was a certain dataset from OpenTorontData that had the same nature of data
(Source: https://open.toronto.ca/dataset/daily-shelter-occupancy/) However, the
data came from 2020 and earlier. The shelter situation from PreCovid was much
more mild and the data from those years would not be an accurate representation
nor an accurate measure of the potentials in the coming years. Therefore, I 
chose the more recent dataset.

## Conlusion
The results show a predominantly adult mix adult demographic in shelters
throughout the GTA. I think one fault of this dataset is its ambiguous sector
labeling as I feel that mixed adults could easily have been separated into the
two sectors of Men and WOmen. However, this dataset does show that most people
who occupies the shelters are adults. Furthermore, the data shows that the
demographics that occupy the shelters seem to be consistent on a day to day 
basis. There does not seem to be huge shifts in people count. I think to notice
a significant difference of shelter population flow, a longer time period must
be analysed. Instead of changes daily, perhaps changes weekly or monthly would
show significant change.

Based on the data, we can posit that more resources must be used to help the
adults and family in these shelters, as they are the predominant demographic. 

## References



